\section{Experimental insights}
This chapter shows effective, model-based regulation with planar \gls{HSA} robots.
The excellent agreement of the model with the actual system behavior (as shown in Sec.~\ref{sec:hsamodel:planar_hsa_robot_model}) enables our model-based controllers to perform very well at the task of setpoint regulation.
For the model-based controllers, any mismatch between the dynamic model and the actual system (as analyzed in Section~\ref{sub:hsacontrol:experiments:model_verification}) has two impacts: (i) the steady-state planning provides us with a desired configuration $q^\mathrm{d}$ which the underactuated robot cannot achieve. This then, in turn, causes a small steady-state error in the end-effector position as seen for the manual setpoints in Fig.~\ref{fig:hsacontrol:experimental_results:fpu:p_sati_d:pee} \& \ref{fig:hsacontrol:experimental_results:fpu:p_sati_d_plus_gc:pee} 
and for the continuous references in Fig.\ref{fig:hsacontrol:experimental_results:fpu:task_space_trajectories}. This steady-state error is absent in the baseline PID as its integral term acts directly in task space. We suggest that future work include an integral term directly on the end-effector position to remove the remaining steady-state error of the model-based controller. Secondly, as (ii), model errors will lead to an offset in the planned steady-state actuation $\phi^\mathrm{ss}$. Therefore, applying a constant $\phi^\mathrm{ss}$ will not move the robot exactly to $p_\mathrm{ee}^\mathrm{d}$. As shown in Fig.~\ref{fig:hsacontrol:experimental_results:fpu:p_sati_d:phi}, the P-satI-D feedback term can compensate for this effect through its proportional and integral terms applied in the collocated variables \& \ref{fig:hsacontrol:experimental_results:fpu:p_sati_d_plus_gc:phi} 
As clearly seen in Fig.~\ref{fig:hsacontrol:experimental_results:fpu:p_sati_d:phi}, the control input drifts away from the planned $\phi^\mathrm{ss}$ to counteract the modeling errors.

In particular, for small to medium bending and axial strains, we observe a very good agreement between the model and the actual system behavior and, therefore, excellent control performance. 
The story slightly changes for more significant bend and twist angles. While the system identification showed that the expressiveness of the model is sufficient, we saw for some of our control experiments significant errors in our feedforward terms, which needed to be compensated by the P-satI-D feedback controller. Precisely, Fig.~\ref{fig:hsacontrol:experimental_results:q} displays errors in the planned shear strain, which subsequently also leads to minor steady-state deviations in the x-coordinate of the end-effector as it can be seen in Fig.~\ref{fig:hsacontrol:experimental_results:pee}. We hypothesize that these errors are mainly caused by the hysteresis characteristics of the \glspl{HSA}~\cite{good2022expanding}. This hypothesis is corroborated by the fact that we had to re-calibrate the axial rest strain $\sigma_\mathrm{ax}^0$ before the start of each experiment.