\chapter*{Curriculum Vit\ae}
\addcontentsline{toc}{chapter}{Curriculum Vit\ae}
\setheader{Curriculum Vit\ae}

%% Print the full name of the author.
\makeatletter
\authors{\@firstname\ {\titleshape\@lastname}}
\makeatother

\noindent
% \begin{longtable}{p{.225\textwidth} p{.70\textwidth}}
%     15/12/1996 & Date of birth in Göppingen, Germany.
% \end{longtable}

\section*{Education}
\noindent
\begin{longtable}{p{.20\textwidth} p{.75\textwidth}}
    03/2021 - 02/2025 & Ph.D. Candidate at the Physical Intelligence (PhI) Lab, Department of Cognitive Robotics, Faculty of Mechanical Engineering, Delft University of Technology, The Netherlands. \newline 
    Supervisor: Prof. Cosimo Della Santina \newline 
    Promotor: Prof. Robert Babuška \newline 
    Research visits: Prof. Daniela Rus, MIT CSAIL, Cambridge, USA (two months); Prof. Gioele Zardini, MIT LIDS, Cambridge, USA (four months). \\
    01/2019 - 02/2021 & Master of Science ETH in Mechanical Engineering, ETH Zürich, Switzerland. Graduated with Distinction. \newline
    Major: Robotics, Systems \& Control
    Tutor: Prof. Emilio Frazzoli \newline
    Exchange Semester: University College London (UCL), UK. \newline
    Thesis: \emph{Solving Occlusion in Digital Elevation Maps Using Neural Networks} as a joint project between RSL at ETH and the PRL at ESA.\\
    09/2014 - 08/2018 & Bachelor of Science ETH in Mechanical Engineering, ETH Zürich, Switzerland. \newline Thesis: Control of all-wheel drives for motorcycles\\
    08/2010 - 06/2014 & Swiss Upper Secondary School Diploma, Kantonsschule Romanshorn, Switzerland. \newline
    Exchange Year: Sentinel Secondary School, West Vancouver, Canada. \newline
    Major: Physics and Applied Mathematics \newline
    Minor: Economics \& Law \newline
    Matura thesis: Adaption to Climate Change in Northern Canada concerning Settlement Structure \newline
\end{longtable}

\section*{Experience}
\noindent
\begin{longtable}{p{.20\textwidth} p{.75\textwidth}}
    Since 01/2021 & Technical Advisor at Quick Technologies AG, Hünenberg, Switzerland.\\
    09/2024 - 12/2024 & Visiting Researcher at the Zardini Lab, Laboratory for Information \& Decision Systems (LIDS), Massachusetts Institute of Technology, Cambridge, USA.\\
    07/2024 - 08/2024 & Visiting Researcher at the Distributed Robotics Lab, Computer Science and Artificial Intelligence Laboratory (CSAIL), Massachusetts Institute of Technology, Cambridge, USA.\\
    07/2024 - 08/2024 & Visiting Researcher at the Planetary Robotics Lab, ESTEC, European Space Agency.\\
    09/2019 - 10/2020 & Team-lead Simulations \& Control at Academic Space Initiative Switzerland (ARIS Space), Zürich, Switzerland.\\
    09/2019 - 10/2020 & Teaching Assistant at IWF, ETH Zürich, Switzerland.\\
    06/2019 - 09/2019 & Research Assistant at Department of Mechanical Engineering, University College London, UK.\\
    09/2018 - 12/2018 & Intern in Product Development at Zühlke Engineering, Schlieren, Switzerland.\\
    02/2018 - 05/2018 & Teaching Assistant at IWF, ETH Zürich, Switzerland.\\
    09/2015 - 12/2020 & Co-Lead Development \& IT at Quap GmbH, Zürich, Switzerland.\\
    03/2016 - 06/2026 & Engineering Intern at Bühler, Uzwil, Switzerland.\\
    08/2014 - 09/2014 & Workshop Trainee at Stadler Rail, Bussnang, Switzerland.\\
    10/2011 - 10/2011 & Student Intern at GFL Consult, Dresden, Germany.
\end{longtable}

\section*{Awards}
\begin{itemize}
    \item[\faTrophy] Named as a \textbf{Rising Star} at the 2025 IEEE International Conference on Soft Robotics (\textbf{RoboSoft 2025}) with a \SI{18}{\percent} selection rate.  
    \item[\faTrophy] RoboSoft 2024 \textbf{Best Paper Award Winner} for publication \faFileTextO \ \emph{M. Stölzle*, S. S. Baberwal*, D. Rus, S. Coyle, and C. Della Santina (2024). Guiding Soft Robots with Motor-Imagery Brain Signals and Impedance Control. In Proceedings of The 2024 IEEE 7th International Conference on Soft Robotics (RoboSoft) (pp. 1-8). IEEE}~\citep{stolzle2024guiding}.
    \item[\faTrophy] Cultuurfonds Wetenschapsbeurzen 2024 supporting six-month research visit at Massachusetts Institute of Technology (MIT), Cambridge, USA\footnote{\url{https://www.cultuurfonds.nl/cultuurfondsbeurzen/wetenschap-beurzen}}.
    \item[\faTrophy] Qualcomm Innovation Fellowship Europe 2024 finalist\footnote{\url{https://www.qualcomm.com/research/university-relations/innovation-fellowship/2024-europe}}.
    \item[\faTrophy] MathWorks poster award at the \href{https://sites.google.com/view/robosoft2023-workshop-rom/home}{RoboSoft 2023 workshop} on \emph{Reduced-Order Modelling for Soft Robots}.
\end{itemize}

\section*{Mentoring}
\noindent
\begin{longtable}{p{.20\textwidth} p{.75\textwidth}}
    Kiwan Wong & Ph.D. student at MIT Mechanical Engineering with Prof. Gioele Zardini. Mentored the student during the visiting period at LIDS with research projects on the co-design of robotic hands and safety-aware control of soft robots using \glspl{CBF} and \glspl{CLF}.\\
    Gabriele di Marzo & Visiting MSc student from Sapienza Università di Roma. Master thesis on \emph{Multistable Motion Primitives}.\\
    Ricardo Valadas & MSc Robotics student at TU Delft. Master thesis on \emph{Low-Dimensional Kinematic and Dynamic Model Identification for Planar Continuum Soft Robots from Image Pixels}. Resulted in a RoboSoft conference paper~\citep{valadas2025learning}\\
    Riccardo Sepe & Visiting MSc student from the University of Turin. Master thesis on \emph{Physics-Informed Model-Based Reinforcement Learning for Soft Robot control}.\\
    BEP 2023 & Supervised a group of four TU Delft BSc Mechanical Engineering students working on \emph{Integration of Planning, Control, and Simulation for Manipulation with a Soft Continuum Robot using Power Grasping}.\\
    BEP \& Applied AI Project 2021 & Supervised a group of four TU Delft BSc Mechanical Engineering students working on \emph{Proprioception for Continuum Soft Robots with Magnetic Sensors}. Resulted in Soft Matter journal paper~\citep{baaij2023learning}.\\
    Emanuele Rosi & Visiting MSc student from the University of Genova. Master thesis on \emph{Sensing soft robots' shape with cameras: an investigation on kinematics-aware SLAM}. Resulted in a RoboSoft conference paper~\citep{rosi2022sensing}.\\
\end{longtable}

\section*{Teaching}
\noindent
\begin{longtable}{p{.23\textwidth} p{.75\textwidth}}
    Intelligent Control \newline Systems 2025 & Onboarding of the new ICS Practical Assignment (PA) TA team.\\
    Intelligent Control \newline Systems 2024 & Lead the creation \& revision of the ICS Practical Assignment (PA), TA for the Q\&A sessions, answered questions on the Brightspace forum, graded the submitted PAs.\\
    Intelligent Control \newline Systems 2023 & Lead the creation of the ICS Practical Assignment (PA), contributed new exercises to the PA (e.g., learning a model with Lagrangian Neural Networks~\citep{lutter2019deep}, Iterative Learning Control, etc.), TA for the Q\&A sessions, answered questions on the Brightspace forum, graded the submitted PAs.\\
    Intelligent Control \newline Systems 2022 & Assisted with the ICS literature assignment (e.g., proposing topics, chairing the symposium, grading submitted assignments).\\
\end{longtable}

\section*{Academic service}
\noindent
\begin{longtable}{p{.20\textwidth} p{.75\textwidth}}
    Organizer & RSS 2024 Workshop on Structural Priors as Inductive Biases for Learning Robot Dynamics\\
    Reviewer & 
        EAAI 2022 \newline
        ERF 2024 EIC Pathfinder Challenge WS \newline
        Humanoids 2024 \newline
        ICRA 2022-2024 \newline 
        IJRR 2023 \newline
        IROS 2023 \newline
        npj robotics 2024 \newline
        RA-L 2021-2024 \newline
        RAM 2022 \newline
        RoboSoft 2022-2025 \newline
        ROPM 2022 \newline
        RSS 2024 Priors4Robots WS \newline
        T-RO 2021-2022
    \\
\end{longtable}

\section*{Other Activities}
\noindent
\begin{longtable}{p{.20\textwidth} p{.75\textwidth}}
    EMERGE & Led the effort and writing of deliverables D3.1 and D3.2 and contributed to deliverable D3.3.\\
\end{longtable}