\section{Conclusion}\label{sec:osmp:conclusion}
In this chapter, we introduced an approach for learning periodic/cyclical motions from demonstrations using \glspl{OSMP}, where—similar to \glspl{DMP}~\citep{ijspeert2002learning, ijspeert2013dynamical}—the motion policy is parametrized by a dynamical system. We combined the advantages of \gls{ML} methods, such as their high expressiveness, with nonlinear system theory to ensure that the periodic motion is tracked via a limit cycle that exhibits \gls{AOS} and transverse contraction. This is achieved by integrating a learned bijective encoder based on Euclideanizing flows~\citep{dinh2016density, rana2020euclideanizing}, which establishes a diffeomorphism between the oracle and latent space, with fixed, inspectable dynamics in latent space provided by a supercritical Hopf bifurcation~\citep{strogatz2018nonlinear}. Compared to previous work~\citep{zhi2024teaching}, our approach significantly improves the accuracy of matching the periodic demonstration to the generated limit cycle. Furthermore, benchmarking against standard \gls{ML} techniques shows that \glspl{OSMP} guarantee convergence, whereas traditional methods such as \glspl{RNN} and \glspl{NODE}~\citep{zhi2024teaching} often exhibit spurious attractors. We also validate our method extensively in real-world settings across various robotic platforms, including robotic manipulators and \glspl{Cobot}, soft robots, and bioinspired hybrid soft-rigid underwater robots.

Moreover, we developed several techniques to enhance the capabilities of dynamic motion primitives. These include online reshaping of the velocity field to adjust convergence characteristics without retraining, synchronizing multiple \glspl{OSMP} in phase via an error-based feedback term, and enabling the same motion policy to learn multiple distinct behaviors by conditioning the encoder on the task. With the addition of a tailored loss term, we can even achieve smooth interpolation between two motion behaviors.

% For future work, it would be intriguing to enable a single motion primitive to exhibit multiple classes of attractors~\citep{strogatz2018nonlinear}, such as \gls{GAS} for point-to-point motions, \gls{MS} for multiple equally valued goals, and \gls{OS} for periodic motions. An initial step could build on the formulation presented here, as the supercritical Hopf bifurcation can capture both a single isolated equilibrium and limit cycle behavior if an additional \emph{attractor type} parameter is introduced. Furthermore, it appears timely to condition the encoder on the outputs or embeddings of modern \glspl{VLM}~\citep{o2024open, touvron2023llama, grattafiori2024llama}, which would significantly enhance the generalization, reasoning, and planning capabilities of the motion policies while preserving the insight, robustness, compliance, and convergence guarantees inherent to \glspl{DMP}.
