\chapter{Modeling of Handed Shearing Auxetics (HSA) Robots}
\label{chp:hsamodel}

\begin{foreword}
    Novel soft robots based on \glspl{HSA} show great promise by integrating multiple \gls{DOF} in a compact form. Their parallel nature promises a larger payload capacity while preserving compliance. Notably, they represent a paradigm shift in the soft robotics domain by i) establishing compliance through their metamaterial structure rather than material softness and ii) transmitting actuation through their elasticity with the auxetic metamaterial instead of directly applying forces and torques into the desired directions of deformation.
    However, despite the extensive literature on their design, fabrication, and integration of sensing, the absence of kinematic and dynamic models hinders their widespread adoption. These models are crucial for developing accurate simulators and effective model-based controllers.
    This chapter addresses this gap by demonstrating how kinematic (Section~\ref{sec:hsamodel:hsa_rod_kinematics}) and dynamical (Section~\ref{sec:hsamodel:hsa_robot_simulation}) models can be derived for general 3D motion scenarios. Furthermore, we present in Section~\ref{sec:hsamodel:planar_hsa_robot_model} low-dimensional kinematical parametrizations and a control-oriented Euler-Lagrangian model tailored explicitly to planar \gls{HSA} robot motions.
    We follow here a strategy of exploiting the vast and well-grounded literature on control-oriented modeling of rigid robotic manipulators and augment them with a specialized kinematic model based on the \gls{PCS} parametrization and a nonlinear, actuation-dependent elastic model.
\end{foreword}

\blfootnote{This chapter is partly based on 
\begin{itemize}
    \item[\faFileTextO] ~\emph{\textbf{M. Stölzle}, L. Chin, R. L. Truby, D. Rus, and C. Della Santina (2023, April). Modeling Handed Shearing Auxetics: Selective Piecewise Constant Strain Kinematics and Dynamic Simulation. In 2023 IEEE International Conference on Soft Robotics (RoboSoft) (pp. 1-8). IEEE}~\cite{stolzle2023modelling}.
    \item[\faFileTextO] ~\emph{\textbf{M. Stölzle}, D. Rus, and C. Della Santina (2023, December). An Experimental Study of Model-based Control for Planar Handed Shearing Auxetics Robots. In Experimental Robotics: The 18th International Symposium. Springer}~\cite{stolzle2024experimental}.
\end{itemize}
}


\pagebreak

\begin{abstract}
    Electrically-actuated continuum soft robots based on \glspl{HSA} promise rapid actuation capabilities while preserving structural compliance. However, the foundational models of these novel actuators required for precise control strategies are missing. This chapter proposes two key components extending the \glsxtrfull{DCM} to allow for modeling \glspl{HSA}. First, we propose a mechanism for incorporating the auxetic trajectory into \gls{DCM} dynamical simulations. We also implement this extension as a plugin for the Elastica simulator. Second, we introduce a \gls{SPCS} kinematic parameterization that can describe an \gls{HSA} rod's shape with fewer configuration variables. 
    Subsequently, based on the augmented \gls{DCM} model, we devise a low-dimensional kinematic and a control-oriented dynamical model for planar \gls{HSA} robots.
    We verify all theoretical contributions experimentally. The \gls{HSA} robot simulator is used to replicate experimental data of the mechanical characterization of HSA rods. For the kinematic description of the \gls{HSA} rods, we attach motion capture markers at various points to a parallel HSA robot and find that the shape of the \glspl{HSA} can be kinematically represented with an average accuracy of \SI{0.3}{mm} for positions and \SI{0.07}{rad} for orientations.
    Finally, the identified Euler-Lagrangian model for planar \gls{HSA} robots exhibits exceptional accuracy at predicting future system states over a horizon of more than \SI{4}{s}.
\end{abstract}

%% Start the actual chapter on a new page.
\newpage

\input{hsamodel/sections/S01_introduction}
\input{hsamodel/sections/S02_hsa_robot_simulation}
\input{hsamodel/sections/S03_hsa_rod_kinematics}
\input{hsamodel/sections/S04_planar_hsa_robot_model}
\input{hsamodel/sections/S05_conclusion}


\section*{Afterword}
This chapter presented approaches for representing the kinematics and dynamics of \gls{HSA} rods in the general case. Additionally, we introduced the first low-dimensional kinematic parameterizations and control-oriented dynamical models for planar \gls{HSA} robots.
Next, Chapter~\circled{\ref{chp:hsacontrol}} will focus on leveraging the models proposed in this chapter for model-based control of planar \gls{HSA} robots.
Namely, this includes the development of stable and effective configuration-space PID+feedforward and task-space impedance controllers.
Subsequently, Chapter~\circled{\ref{chp:braincontrol}} will develop a \gls{BCI} strategy for guiding soft robots, and specifically \gls{HSA} robots, using motor imagery in a safe and compliant fashion.