% \documentclass[conference]{IEEEtran}
% \documentclass[letterpaper, 10 pt, conference]{ieeeconf}  % Comment this line out if you need a4paper
\documentclass[a4paper, 10pt, conference]{ieeeconf}      % Use this line for a4 paper

\IEEEoverridecommandlockouts                              % This command is only needed if 
                                                          % you want to use the \thanks command

\overrideIEEEmargins                                      % Needed to meet printer requirements.

\usepackage{cite}

% math
\let\proof\relax
\let\endproof\relax
\usepackage{amsthm, amsmath,amssymb,amsfonts}
\usepackage{cases}

\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage{siunitx}

\usepackage{booktabs}       % professional-quality tables
\usepackage{tabularx}

% figures
\usepackage[font=small]{caption}
\usepackage{epsfig} % for postscript graphics files
\usepackage{svg}
\usepackage{tikz}
% we can either use subcaption or subfigure
\usepackage{subfigure}
% \usepackage{subcaption}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
    
% math proof
\theoremstyle{plain}
\newtheorem{theo}{Theorem}
\newtheorem{Remark}{Remark}
\newtheorem{Lemma}{Lemma}
\newtheorem{Proposition}{Proposition}
\newtheorem{Definition}{Definition}
\newtheorem{Corollary}{Corollary}
\newtheorem{exmp}{Example}%[section]


% Space modifications
\setlength{\textfloatsep}{3pt}
\setlength{\abovedisplayskip}{3pt}
\setlength{\belowdisplayskip}{3pt}
\setlength{\abovedisplayshortskip}{3pt}
\setlength{\belowdisplayshortskip}{3pt}

\renewcommand{\baselinestretch}{.99} 

% THEOREMS
\DeclareMathOperator*{\argmin}{\arg\!\min}
\DeclareMathOperator*{\argmax}{\arg\!\max}

% algorithms
\usepackage{algorithm}
\usepackage{algorithmic}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}

\usepackage{acro}
\newcommand{\newac}[2]{\DeclareAcronym{#1}{short=#1,long=#2}}
\newac{BA}{Bundle Adjustment}
\newac{CAD}    {Computer-aided Design}
\newac{CC}    {Constant Curvature}
\newac{DoF}    {Degrees of Freedom}
\newac{MoCap}    {Motion-capture}
\newac{MSE}    {Mean Squared Error}
\newac{PCC}    {Piecewise Constant Curvature}
\newac{PCS}    {Piecewise Constant Strain}
\newac{RMSE}    {Root-Mean-Square Error }
\newac{SLAM}    {Simultaneous Localization and Mapping}
\newac{SVR}    {Support Vector Regression}
\newac{vSLAM}   {visual SLAM}
    
    
\begin{document}
% shorten the author list to et al. if more than 2
% This will limit author lists more than 3 by printing only the first 2 authors followed by et al.
% https://tex.stackexchange.com/questions/164017/limiting-the-number-of-authors-in-the-references-with-ieeetran
\bstctlcite{IEEEexample:BSTcontrol}

% \addtolength{\topmargin}{50pt}

%\title{Pose Estimation for a soft robotic segment through SLAM
%
%\thanks{Identify applicable funding agency here. If none, delete this.}
%}
\title{\LARGE \bf
Sensing soft robots' shape with cameras: \\ an investigation on kinematics-aware SLAM
}

% Full author description in header
% \author{\IEEEauthorblockN{1\textsuperscript{st} Maximilian Stölzle}
% \IEEEauthorblockA{\textit{Cognitive Robotics} \\
% \textit{Delft University of Technology}\\
% Delft, Netherlands \\
% M.W.Stolzle@tudelft.nl}
% \and
% \IEEEauthorblockN{2\textsuperscript{nd} Cosimo Della Santina}
% \IEEEauthorblockA{\textit{Cognitive Robotics} \\
% \textit{Delft University of Technology}\\
% Delft, Netherlands \\
% C.DellaSantina@tudelft.nl}
% }

% Short variant
\author{Emanuele Riccardo Rosi$^{*, 1, 2}$, Maximilian Stölzle$^{*, 1}$, Fabio Solari$^{2}$, Cosimo Della Santina$^{1,3}$ % <-this % stops a space
\thanks{%
$^*$Authors contributed equally.
$^{1}$Cognitive Robotics department, Delft University of Technology, Mekelweg 2, 2628 CD Delft, Netherlands. {\tt\scriptsize \{M.W.Stolzle, C.DellaSantina\}@tudelft.nl} %
$^{2}$Department of Computer Science, Bioengineering, Robotics and System Engineering (DIBRIS), University of Genoa, Viale Causa 13, 16145 Genoa, Italy {\tt\scriptsize s4361996@studenti.unige.it, fabio.solari@unige.it} %
$^{3}$ Institute of Robotics and Mechatronics, German Aerospace Center (DLR), 82234 Weßling, Germany
}%
}

\maketitle

\begin{abstract}
%
The nature of continuum soft robots calls for novel perception solutions, which can provide information on the robot's shape while not substantially modifying their bodies' softness. One way to achieve this goal is to develop innovative and completely deformable sensors. However, these solutions tend to be less reliable than classic sensors for rigid robots. As an alternative, we consider here the use of monocular cameras. By admitting a small rigid component in our design, we can leverage well-established solutions from mobile robotics. We propose a shape sensing strategy that combines a SLAM algorithm with nonlinear optimization based on the robot's kinematic model. We prove the method's effectiveness in simulation and with experiments of a single-segment continuous soft robot with a camera mounted to the tip.
We achieve mean relative translational errors below 9\% simulations and experiments alike, and as low as 0.5\% on average for some simulation conditions.
%
\end{abstract}

\acresetall

\input{sections/S01_introduction}

%\input{sections/S02_related_work}

\input{sections/S03_pose_estimation}

\input{sections/S04_simulations}

\input{sections/S05_experiments}

\input{sections/S06_conclusion}

\bibliographystyle{IEEEtran}
\bibliography{bibliography/bibliography}

\end{document}
