\chapter*{List of Publications}
\addcontentsline{toc}{chapter}{List of Publications}
\setheader{List of Publications}
\label{publications}

\section*{Referred journals}

%% We use the 'etaremune' environment (the reverse of 'enumerate') to get a
%% numbered list of publications in reverse chronological order. If the list of
%% authors is long, it might be useful to emphasize your own name with \textbf.
% APA reference style with modifications
\begin{enumerate}{
    \item[\faFileTextO \, \stepcounter{enumi}\arabic{enumi}.] \textbf{M. Stölzle}, T.K. Rusch*, Z.J. Patterson*, R. Pérez Dattari, F. Stella, J. Hughes, C. Della Santina, and D. Rus (2025). Learning Stable Periodic Robotic Motions from Demonstration. In Science Robotics, \textbf{\emph{In Preparation}}.
    \item[\faFileTextO \, \stepcounter{enumi}\arabic{enumi}.] \textbf{M. Stölzle}*, N. Pagliarani*, F. Stella*, J. Hughes, C. Laschi, D. Rus, M. Cianchetti, C. Della Santina, and G. Zardini (2025). Safe yet Effective Soft Robots via Holistic Co-Design. In Nature Machine Intelligence, \textbf{\emph{In Preparation}}.
    \item X. Shao, P. Pustina*, \textbf{M. Stölzle}*, G. Sun, A. De Luca, L. Wu, and C. Della Santina (2023). Model-based control for soft robots with system uncertainties and input saturation. IEEE Transactions on Industrial Electronics, 71(7), 7435-7444.
    \item[\faFileTextO \, \stepcounter{enumi}\arabic{enumi}.] T. Baaij*, M. K. Holkenborg*, \textbf{M. Stölzle}*, D. van der Tuin*, J. Naaktgeboren, R. Babuška, and C. Della Santina (2023). Learning 3D shape proprioception for continuum soft robots with multiple magnetic sensors. Soft Matter, 19(1), 44-56.
    \item \textbf{M. Stölzle}, T. Miki, L. Gerdes, M. Azkarate, and M. Hutter (2022). Reconstructing occluded elevation information in terrain maps with self-supervised learning. IEEE Robotics and Automation Letters, 7(2), 1697-1704.
    \item[\faFileTextO \, \stepcounter{enumi}\arabic{enumi}.] \textbf{M. Stölzle}, C. Della Santina (2021). Piston-driven pneumatically-actuated soft robots: Modeling and backstepping control. IEEE Control Systems Letters, 6, 1837-1842.
}\end{enumerate}


\section*{Referred conference proceedings}

%% We use the 'etaremune' environment (the reverse of 'enumerate') to get a
%% numbered list of publications in reverse chronological order. If the list of
%% authors is long, it might be useful to emphasize your own name with \textbf.
% APA reference style with modifications
\begin{enumerate}
    \item[\faFileTextO \, \stepcounter{enumi}\arabic{enumi}.] R. Valadas*, \textbf{M. Stölzle}*, J. Liu, and C. Della Santina (2025). Learning Low-Dimensional Strain Models of Soft Robots by Looking at the Evolution of Their Shape with Application to Model-Based Control. In Proceedings of the 2025 IEEE 8th International Conference on Soft Robotics (RoboSoft) (pp. 1-8). IEEE. \emph{Accepted}.
    \item[\faFileTextO \, \stepcounter{enumi}\arabic{enumi}.] \textbf{M. Stölzle}, and C. Della Santina (2024). Input-to-State Stable Coupled Oscillator Networks for Closed-form Model-based Control in Latent Space. In Proceedings of Advances in Neural Information Processing Systems (NeurIPS) 37, \textbf{Spotlight (top 10\% of accepted papers)}.
    % \item[\stepcounter{enumi}\arabic{enumi}.] G. Buriani*, J. Liu*,  \textbf{M. Stölzle}, C. Della Santina, and J. Ding (2024). Symbolic Learning of Interpretable Reduced-order Models for Jumping Legged Robots. In Proceedings of IEEE-RAS International Conference on Humanoid Robots. IEEE, \emph{Under Review}.
    \item[\faFileTextO \, \faTrophy \, \stepcounter{enumi}\arabic{enumi}.] \textbf{M. Stölzle}*, S. S. Baberwal*, D. Rus, S. Coyle, and C. Della Santina (2024). Guiding Soft Robots with Motor-Imagery Brain Signals and Impedance Control. In Proceedings of the 2024 IEEE 7th International Conference on Soft Robotics (RoboSoft) (pp. 1-8). IEEE. \textbf{Best Paper Award}.
    \item[\faFileTextO \, \stepcounter{enumi}\arabic{enumi}.] \textbf{M. Stölzle}, D. Rus, and C. Della Santina (2024). An Experimental Study of Model-based Control for Planar Handed Shearing Auxetics Robots. In Experimental Robotics: The 18th International Symposium (ISER) (pp. 153-167). Springer.
    \item A. Ceni*, A. Cossu*, \textbf{M. Stölzle}, J. Liu, C. Della Santina, D. Bacciu, and C. Gallicchio (2024). Random Oscillators Network for Time Series Processing. In Proceedings of The 27th International Conference on Artificial Intelligence and Statistics (AISTATS) (pp. 4807-4815). PMLR.
    \item[\faFileTextO \, \stepcounter{enumi}\arabic{enumi}.] \textbf{M. Stölzle}, L. Chin, R. L. Truby, D. Rus, and C. Della Santina (2023, April). Modelling Handed Shearing Auxetics: Selective Piecewise Constant Strain Kinematics and Dynamic Simulation. In 2023 IEEE International Conference on Soft Robotics (RoboSoft) (pp. 1-8). IEEE.
    \item[\faFileTextO \, \stepcounter{enumi}\arabic{enumi}.] E. Rosi*, \textbf{M. Stölzle}*, F. Solari, and C. Della Santina (2022, April). Sensing Soft Robots' Shape with Cameras: an Investigation on Kinematics-Aware SLAM. In 2022 IEEE 5th International Conference on Soft Robotics (RoboSoft) (pp. 795-801). IEEE.
    \item T. Phillips*, \textbf{M. Stölzle}*, E. Turricelli*, F. Achermann, N. Lawrance, R. Siegwart, and J. J. Chu (2021, May). Learn to Path: Using Neural Networks to Predict Dubins Path Characteristics for Aerial Vehicles in Wind. In 2021 IEEE International Conference on Robotics and Automation (ICRA) (pp. 1073-1079). IEEE.
\end{enumerate}

\vspace{0.5cm}
\noindent
\faFileTextO~~Included in this thesis.\\
*~~indicates equal contribution.\\
\faTrophy~~Won a best paper award.

\section*{Invited talks \& dissemination}
\noindent
\begin{longtable}{p{.125\textwidth} p{.85\textwidth}}
    04/2025 & \emph{Safe Control of Soft Robots: Bridging Physics and Learned Models} at 2025 IEEE International Conference on Soft Robotics (RoboSoft 2025), Lausanne, Switzerland. Rising Stars talk.\\
    04/12/2024 & \emph{Model-based Control of Planar Handed Shearing Auxetics Robots} at MRS Fall Meeting, Boston, USA. Contributed talk.\\
    01/10/2024 & \emph{Integrating Physical Structure and Stability Guarantees into the Learning of Robot Models \& Motion Policies} at MIT LIDS Autonomy Tea Talk, Cambridge, USA. Invited talk.\\
    12/09/2024 & \emph{Designing Compact Models for the Control of Soft Robots} at MIT Zardini Lab Group Meeting, Cambridge, USA. Invited talk.\\
    20/08/2024 & \emph{Deriving and Learning Minimalist Models for the Control of Soft Robots} at MIT Distributed Robotics Lab (DRL) Group Meeting, Cambridge, USA. Invited talk.\\
    15/07/2024 & \emph{Leveraging Coupled Oscillator Networks as a Structural Prior when Learning Latent Dynamics from Pixels} at RSS 2024 Workshop on Structural Priors as Inductive Biases for
    Learning Robot Dynamics, Delft, Netherlands. Spotlight talk (oral). \SI{15}{\percent} acceptance rate (3/27), peer-reviewed, \textbf{rated by reviewers as the best submission}.\\
    25/06/2024 & \emph{Input-to-State Stable Coupled Oscillator Networks for Closed-form Model-based Control in Latent Space} at TU Delft AI PhD Poster Day 2024, Delft, Netherlands. Poster presentation.\\
    05/2024 & \emph{Model-based Control of Planar Handed Shearing Auxetics Robots} at Dutch Soft Robotics Symposium 2024, Eindhoven, Netherlands. Invited talk.\\
    05/2023 & \emph{Modelling Handed Shearing Auxetics: Selective Piecewise Constant Strain Kinematics and Dynamic Simulation} at Dutch Soft Robotics Symposium 2023, Enschede, Netherlands. Poster presentation.\\
    04/2023 & \emph{Learning 3D Shape Proprioception for Continuum Soft Robots with Multiple Magnetic Sensors} at RoboSoft 2023 Workshop on Reduced-Order Modelling for Soft Robots, Singapore. Poster presentation. \textbf{Poster presentation award (3rd-place).}\\
    04/2023 & \emph{Learning 3D Shape Proprioception for Continuum Soft Robots with Multiple Magnetic Sensors} at 3rd International Conference on Embodied Intelligence, Online. Contributed talk.\\
    %
    09/2022 & \emph{Learning 3D Shape Proprioception for Continuum Soft Robots with Multiple Magnetic Sensors} at 15th International Workshop on Human-Friendly Robotics (HFR), Delft, Netherlands. Poster presentation.\\
    05/2022 & \emph{Learning 3D Shape Proprioception for Continuum Soft Robots with Multiple Magnetic Sensors.} at ICRA 2022 Workshop on Compliant Manipulation, Philadelphia, USA. Poster presentation.\\
    04/2022 & \emph{Learning 3D Shape Proprioception for Continuum Soft Robots with Multiple Magnetic Sensors} at RoboSoft 2022 Workshop on Soft Sensing: Environment, Morphology, Brain in Biology and Robotics, Edinburgh, United Kingdom. Poster presentation.\\
    04/2022 & \emph{Piston-Driven Pneumatically-Actuated Soft Robots: modeling and backstepping control} at RoboSoft 2022 Workshop on Soft Robotics modeling: what are we missing?, Edinburgh, United Kingdom. Poster presentation.\\
    12/2021 & \emph{Solving Occlusion in Terrain Mapping with Neural Networks} at NeurIPS 2021 4th Robot Learning Workshop: Self-Supervised and Lifelong Learning, Online. Poster presentation.\\
    05/2021 & \emph{Piston-Driven Pneumatically-Actuated Soft Robots: modeling and backstepping control} at Benelux Workshop on Systems and Control 2021, Rotterdam, Netherlands. Contributed talk.\\
\end{longtable}